{\TheTitle}\thanks{Submitted to the editors DATE. \funding{This work was funded by the Fog Research Institute under contract no.~FRI-454.}}

### 2017-12-19 ssc-0-0.tex
complete second draft of SI paper

% REQUIRED
\begin{abstract}
  This is an example SIAM \LaTeX\ article. This can be used as a
  template for new articles.  Abstracts must be able to stand alone
  and so cannot contain citations to the paper's references,
  equations, etc.  An abstract must consist of a single paragraph and
  be concise. Because of online formatting, abstracts must appear as
  plain as possible. Any equations should be inline.
\end{abstract}

@@

% formula equations
The semi-implicit formula is a rule for integrating systems of
ordinary differential equations \cite{jkp-132-312,jkp-93-144}.  It
is most often used as a method of lines, to solve stochastic partial
differential equations that occur when phase space Monte-Carlo
methods are used in quantum optics and in atom optics \cite{pra-58-4824}.
These equations resemble Schr\"odinger's equation,
\begin{equation}\label{eq:vi}
    \frac{\partial\psi}{\partial t} = -i(-\nabla^2 + V)\psi,
\end{equation}
with additional nonlinear and random terms.  The spatial axes are
discretised, often using spectral derivative operators, to give a
system of ordinary or stochastic differential equations
\cite{2000-Trefethen-Spectral}.  Then these equations are integrated
by the semi-implicit formula.

@@

If the formula of Equation~\cref{eq:i} could be applied directly
to Equation~\cref{eq:iii}, the computed solution would be very
stable.  The coefficient vector~$s_n$ can be expanded over the
eigenvectors~$v_k$ of~$H$, as $s_n=\sum_k c^kv_k$.  \Cref{fig:iii}
shows the trajectory over a time step from~$s_n$ to~$s_{n+1}$,
projected onto the complex plane spanned by the eigenvector~$v_k$.
The exact solution that satisfies Equation~\cref{eq:iii}, with
initial condition~$s_n$, is~$s(h)=\sum_k c^ke^{-i\omega_kh}v_k$,
and its projection lies on the circle around the origin of
radius~$|c_k|$.

@@

{\bf Check growth rates.}

@@

@@

@@

\section*{Acknowledgments}
I would like to acknowledge the support of Swinburne University of Technology.  Thanks to Run Yan Teh for suggestions that improved the presentation.

>>>>

% REQUIRED
\begin{abstract}
The iterative semi-implicit formula, or some variant of it, is often
used to solve stochastic partial differential equations that arise
in quantum optics and in atom optics.  When this formula is applied
to a quantum system with unitary dynamics, stability depends on a
kind of Nyquist condition, where the angular frequency corresponding
to the largest eigenvalue of the discretised hamiltonian is at most
twice the reciprocal time step.  The formula develops parasitic
solutions very abruptly as this condition is exceeded, which makes
it troublesome for practical use, but quite useful as a test of the
stability of discretisation methods.
\end{abstract}

@@

% formula equations
The semi-implicit formula is a rule for integrating systems of
ordinary differential equations \cite{jkp-132-312,jkp-93-144}.  It
is most often used as a method of lines, to solve stochastic partial
differential equations that occur when phase space Monte-Carlo
methods are used in quantum optics and in atom optics \cite{pra-58-4824}.
These equations resemble Schr\"odinger's equation,
\begin{equation}\label{eq:vi}
    \frac{\partial\psi}{\partial t} = -i(-\nabla^2 + V)\psi,
\end{equation}
with additional nonlinear and random terms.  The spatial axes are
discretised, often using spectral derivative operators, to give a
system of ordinary or stochastic differential equations
\cite{2000-Trefethen-Spectral}.  These equations are integrated
by the semi-implicit formula.

@@

If the formula of Equation~\cref{eq:i} could be applied directly
to Equation~\cref{eq:iii}, the computed solution would be very
stable.  The coefficient vector~$s_n$ can be expanded over the
eigenvectors~$v_k$ of~$H$, as $s_n=\sum_k c^kv_k$.  \Cref{fig:iii}
shows the trajectory over a time step from~$s_n$ to~$s_{n+1}$,
projected onto the complex plane spanned by the eigenvector~$v_k$.
The exact solution that satisfies Equation~\cref{eq:iii}, with
initial condition~$s_n$, is~$s(h)=\sum_k c^ke^{-i\omega_kh}v_k$,
and its projection lies on the circle around the origin of
radius~$|c_k|$.  Here~$\omega_k$ is the eigenvalue of~$H$ corresponding to the eigenvector~$v_k$.

@@

@@

\Cref{fig:a} shows how abruptly the semi-implicit formula becomes
unstable as the frequency~$\omega_k$ increases through the limit
of stability.  While this might be unfortunate from a practical
point of view, it makes the formula useful for evaluating new ways
to discretise the hamiltonian.  The largest~$\omega_k$ can be
estimated by integrating the dynamics over a few time steps, and
adjusting the time step to find the limit of stability.  Quantum
states inspire some very clever methods of discretisation
\cite{jcp-64-63,jcp-119-1289}, because the state of a quantum system
with many degrees of freedom lies in a vector space of just as many
dimensions, a space where any grid will have impossibly many points.
It is not always straightforward to estimate the eigenvalues of the
discrete hamiltonians that result from these methods.  In particular,
the discrete hamiltonian can be a non-normal operator, in which
case quite delicate arguments about pseudospectra would be required
to rigorously determine the stability of the discretised dynamics.
Running those dynamics, and looking out for parasitic solutions,
is a fast and simple alternative.

@@

When the equation being solved is simply Schr\"odinger's equation
with repulsion, there are ways to modify the semi-implicit method
and avoid the large energies.  The most commonly used is an interaction
picture method \cite{2000-CaradocDavies-vortex}.  This is a split
step method, where a Fourier transform~$\psi(t,k)$ is taken at each
time step, and the kinetic energy term proportional to~$-i\nabla^2\psi$
is evaluated exactly as~$\psi(t+h,k)=e^{ik^2h}\psi(t,k)$.  This
ensures that~$\|\psi(t+h)\|=\|\psi(t)\|$, so that parasitic modes
can not grow only because of their high kinetic energy.

Interaction picture methods are very effective when they work.
However, the nonlinear terms can also become large, and this occurs
quite often in practice.  For example, it is often desirable to
simulate a cold gas in a harmonic trap, with~$V(x)\propto x^2$.  In
this case, the energy of the gas oscillates between potential and
kinetic, and the potential energy of a particle far from the origin
can be as great as the kinetic energy of a particle with large
momentum.  This causes the equation to remain just as stiff even
when the kinetic energy term is solved exactly.

@@

\section*{Acknowledgments}
I would like to acknowledge the support of Swinburne University of Technology.  Thanks to Run Yan Teh for useful suggestions.
